---
title: "Survey Weighting Exercise: Nonvoter Study Analysis"
date: "2025-10-29"
author: "Derek Willis/Rob Wells"
---

Carla Espinoza-Pinto



Activity: Belonging Survey Methods

Let's revisit the summary of UMD's Belonging Survey, and look at the appendix in particular. https://president.umd.edu/internal/belonging

Working in pairs or small groups, each person reads the appendix closely and identifies any statements that raise methodological concerns or are unclear. Discuss your findings, summarize your discussion and then submit no more than three most important concerns or questions you have and how they might impact the results of the Belonging Survey.


The first concern that could impact the results of the Belonging Survey is that it is unclear how the participants are being picked and if the whole UMD community was exposed to the survey or not. This could impact the results because there could be certain groups or minorities not represented in the survey because the survey wasn't advertised for them. 
The second concern that could impact the results of the Belonging Survey is the fact that the survey results don't seem to match up with UMD's population. This connects with the first concern because if the results were made to look like they represented the whole community of UMD but don't then it appears like certain group's opinions are being favored over others.

The third concern that could impact the results of the Belonging Survey is that there isn't an indication as to who actually responded to the surveys. The numbers, such as how many people were exposed to the survey and how many actually took it are unclear and this causes an issue because we don't know who exactly we are basing these conclusions off of.

#-------------------------------------------------------------------------------
#Survey Weighting
#-------------------------------------------------------------------------------


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)

```

# Introduction

This notebook explores how survey weighting affects data analysis and interpretation in journalism. We'll use a dataset from a 2020 survey on voting behavior to understand weighting concepts and how they can change our understanding of public opinion data. The codebook with the questions from the survey is here: https://github.com/dwillis/jour405_files/blob/main/nonvoters_codebook.pdf

## What is Survey Weighting?

Survey weighting is like adjusting the volume on different voices in a conversation to make sure everyone is heard equally. Think of it like running a mixing board at a concert. In survey research:

- Some groups of people are more likely to answer surveys than others
- This can make the survey sample different from the actual population
- Survey weights help fix this imbalance by giving more "volume" to underrepresented groups and less to overrepresented groups

**For example:** If your survey has too many college graduates compared to the general population, you might give each college graduate's response a weight of 0.8 (counting as 80% of a response) and each non-college graduate a weight of 1.2 (counting as 120% of a response).

## Why Should Journalists Care?

- Unweighted results might give a misleading picture of public opinion
- Proper weighting helps ensure your reporting accurately represents the community
- Understanding weighting helps you critically evaluate polls from other sources
- Small differences in weighting can sometimes change the headline of your story

## Our Questions

1. How does weighting affect our estimates of voting intentions?
2. Which demographic groups show the largest differences between weighted and unweighted results?
3. How might weighting influence reporting of this data?

# Data Overview

The nonvoters dataset contains survey responses about voting intentions and behaviors from 5,836 American adults. The survey was conducted prior to the 2020 election and includes a variety of demographic information and political attitudes.

```{r load-data}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Look at the structure of the data
glimpse(nonvoters_data)
```

Let's break down the key variables we'll focus on:

- `weight`: The survey weight assigned to each respondent, showing how much their response "counts"
- `Q21`: Voting intention with these values:
  - 1 = Yes (plans to vote)
  - 2 = No (does not plan to vote)
  - 3 = Unsure/Undecided
- `voter_category`: Respondent's voting history:
  - "always": Votes in all or nearly all elections
  - "sporadic": Votes occasionally
  - "rarely/never": Rarely or never votes
- Demographics: `race`, `gender`, `educ` (education level), `income_cat`

# Understanding Survey Weights

Before we dive into the analysis, let's examine the weights in our dataset to understand how they're distributed.

```{r examine-weights}
# Summary statistics of weights
summary(nonvoters_data$weight)
```


```{r}
# Histogram of weights
ggplot(nonvoters_data, aes(x = weight)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Survey Weights",
    x = "Weight",
    y = "Count"
  )
```

**Explanation:** 
- The summary statistics show us the minimum, maximum, and typical weights in the dataset
- The histogram shows how many respondents have each weight value
- Most weights cluster around 1.0 (which means those respondents count as exactly one person)
- Weights below 1.0 mean a respondent is overrepresented (counts as less than one person)
- Weights above 1.0 mean a respondent is underrepresented (counts as more than one person)

**Questions:**

1. What is the range of weights in this dataset? 

0.2 to 3.0

2. What does it mean when a respondent has a weight of 0.23? What about 3.04? 

a weight of 0.23 means the respondent is overrepresented
a weight of 3.04 means the respondent is underrepresented

3. Why might some respondents be weighted more heavily than others? *

Certain groups such as young people, low-income families or people with little to no education tend to not be exposed to surveys as often as other groups.

# Exercise 1: Voter Categories and Weighting

First, let's look at how many people fall into each voting category in our sample.

```{r voter-categories}
library(knitr)
# Count of respondents by voter category
nonvoters_data |>
  count(voter_category) |>
  mutate(percentage = n / sum(n) * 100)
```

**Explanation:** This table shows the breakdown of our sample by voting history. We can see that most respondents (44.1%) are "sporadic" voters, followed by "always" voters (31.0%), and "rarely/never" voters (24.9%).

Now, let's compare unweighted and weighted voting intentions by voter category.

## Part A: Unweighted Analysis

First, we'll calculate the raw percentages without applying weights:

```{r unweighted-by-category}
# Calculate unweighted percentages
unweighted_by_voter_category <- nonvoters_data |>
  # Filter out missing responses
  filter(!is.na(Q21), Q21 > 0) |>
  # Group by voter category and response
  group_by(voter_category, Q21) |>
  # Count responses in each group
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages within each voter category
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable version with voting intentions as columns
unweighted_summary <- unweighted_by_voter_category |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

unweighted_summary
```

**Explanation:**
- This table shows the percentage of respondents in each voter category who said "Yes," "No," or "Unsure" when asked if they plan to vote
- We calculated these percentages by simply counting responses and dividing by the total in each category
- No weights have been applied yet—each respondent counts equally
- Notice that even among "rarely/never" voters, a high percentage say they plan to vote in the upcoming election

## Part B: Weighted Analysis

Now, let's apply the survey weights to the same analysis. Instead of just counting respondents, we'll sum their weights:

```{r weighted-by-category}
# YOUR TASK: Calculate the weighted percentages
# Use the weight variable to adjust the counts

weighted_by_voter_category <- nonvoters_data |>
  # Filter out missing responses, same as before
  filter(!is.na(Q21), Q21 > 0) |>
  # Group by voter category and response
  group_by(voter_category, Q21) |>
  # Instead of counting, sum the weights
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  # Calculate percentages based on weighted counts
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

weighted_summary <- weighted_by_voter_category |>
  pivot_wider(
    id_cols = voter_category,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

weighted_summary
```

**Explanation:**
- Now we've calculated percentages by summing the weights instead of simply counting people
- Each respondent contributes to the total based on their assigned weight value
- This gives us estimates that should better reflect the true population
- Notice how some percentages have changed from the unweighted analysis

## Part C: Comparing Differences

Let's create a comparison table to clearly see the differences between weighted and unweighted results:

```{r category-comparison}
comparison <- unweighted_summary |>
  # Join the two tables to compare them
  inner_join(weighted_summary, by = "voter_category", suffix = c("_unweighted", "_weighted")) |>
  # Calculate the differences in percentage points
  mutate(
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the differences for our table
  select(voter_category, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

comparison 
```

**Explanation:**
- This table shows how many percentage points the results changed after applying weights
- Positive numbers mean the weighted percentage is higher; negative numbers mean it's lower
- For example, after weighting, the percentage of "rarely/never" voters saying "Yes" decreased by 3.6 percentage points
- These differences might seem small, but in a close election, even small shifts can be meaningful for reporting

**Answer These Questions:**

1. Which voter category shows the biggest difference between weighted and unweighted results? 

The rarely/never voters had the biggest difference because in the weighted results the perecentage of voters saying "Yes" decreased by 3.7 percentage points.

2. How might these differences affect reporting on likely voter turnout? 

These differences might affect reporting on likely voter turnout because the two datas give off a different interpretation. The unweighted data gives off the impression that more people plan to vote than the actual number of voters. The weighted data is a more accurate representation.

3. Why do you think these differences exist? What does this tell us about who might be over or under-represented in the sample? 

These differences exist because the number of people who actually participate in surveys can not be used as a representation of the population as a whole. The people who don't vote could be overrepresented so weighting balances it out towards more accurate and properly represented data.
