---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
head(pg_crime)
glimpse(pg_crime)
summary(pg_crime)
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

If we compare the data of autos stolen versus the autos that were stolen and recovered, the numbers are drastically different and there way more cars being stolen than being stolen and returned. I feel like these are the categories that most stood out to me becasue of how different they are from eachother yet it's relatively the same crime. These two categories do make the job of summarizing the data more difficult because it's quite odd that most cars that are stolen are not returned to their original owner. This could make it difficult to summarize the data of how many cars are being stolen because most of the cars that are being stolen are never returned so what does that say about our crime rates?

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
library(tidyverse)

pg_crime_averages <- pg_crime %>%
  mutate(

    total_incidents = rowSums(select(., Jan:Dec), na.rm = TRUE),

    average_incidents = total_incidents / 12
  )


```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The category with the highest average number of incidents per month is Accident with 481.33%. The category with the lowest number of incidents per month was Roberry residential with 2.33%. It's interesting to see in which category most of the crime lie on the chart and where least of it lies. The data tells me that there is a lot of variance of crime in Prince George's County. It doesn't surprise me that theft has a high average number of incidents because of the constant campus alerts sent out. But, it is surprising how low the sex offense crime rates are.

### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
library(tidyverse)

summer_crime_averages <- pg_crime_averages %>% 
  mutate(
    june_change  = ((Jun  - average_incidents) / average_incidents) * 100,
    july_change  = ((Jul  - average_incidents) / average_incidents) * 100,
    august_change = ((Aug  - average_incidents) / average_incidents) * 100
  )

```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

Assault shooting went from a change of -4% in June to a change of 34% in July. There are some patterns in the data, but there is also a large amount of variance so it's difficult to find a correlation between the data. Commercial Robberies drastically went up in July, so there could be some kind of correlation between the month of July that makes it more susceptible to robberies than in June.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Crime data in Prince George's County suggests correlation between summer months and a change of average number of incidents. 

Data collected from Prince George's County crime rates demonstrate that commercial robberies had a percentage change of -0% in July but it had a spike in August resulting in a percentage change of 43%. 

On the contrary, vehicle robberies dropped from a percentage change of 64% in July to a percentage change of -15% in August.

This data suggests that businesses are more susceptible to robberies in the month of August, which could be for a variety of reasons inluding the chaos from back to school shopping, the weather, etc.

The data also suggestst that vehicles are more susceptible to robberies in the month of July, when Americans usually travel, spend more time outdoors and are preparing for fourth of July.

Both data differs in conclusions but observing the crime trends for each month could benefit the community to be more alert. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
