---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv(
  "https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv"
)

```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

There is a variety of salaries, giving us a good amount of data to report on. What stands out to me are the  employees with a salary represented as a 0 because as a reporter you question why their salary may be disclosed. These values of 0 will certainly cause problems for analysis because there is no data.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries %>%
  summarize(
    mean_salary   = mean(salary, na.rm = TRUE),
    median_salary = median(salary, na.rm = TRUE)
  )
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries %>%
  summarize(
    mean_salary   = mean(salary, na.rm = TRUE),
    median_salary = median(salary, na.rm = TRUE),
    min_salary    = min(salary, na.rm = TRUE),
    max_salary    = max(salary, na.rm = TRUE)
  )
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and the median are approximately the same which provides us with a sense of where all these salaries meet. I think that the zeros on this data create outliers in the data but having a median and mean in the 100k range sounds just about right, because most of these employees are either making over 100k or getting there. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I would like to expand on the data further since there are so many salaries being analyzed, so I would break the data down by status. Find out more about what are the qualifications for an employee versus a detailee and compare their salaries. I would also dive into each title and find out which employee under that title has a higher salary and why.
